Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-03-28 manuscript/Chapter2-manuscript.Rmd
updated MS version names

Hunting experience is predicted to increase predator foraging specialization when resources are predictable [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, when resources vary, others predict that learning with experience should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022]. Our results show that both mechanisms operated simultaneously, such that there was no directional increase in behavioural specialization or plasticity in the population. Instead, the consequences from experience appear to be a property of individuals, such that each predator developed, through extensive practice, their own expertise and behavioural trajectory; some specialized at fast-paced hunting, while others adopted a slower and more flexible tactic. The predators' tactics were matched to prey encounters, such that specialized cursorial hunters experienced more predictable encounters, while slower and flexible hunters experienced more unpredictable encounters, with no difference in success between the two predator-types. This predator-prey behavioural matching grew stronger with predator experience, while the relationship between the prey's behaviour and hunting success diminished. Together, our observations suggest that predators adjust their hunting tactic with experience, and the degree at which they specialize in it, to the behaviour of the prey that they encountered throughout their lifetime in the game. Moreover, the predator-prey behavioural escalation through experience suggests that learning may be a catalyst for arms races.

@@

Empirical research in humans and animals shows that task proficiency often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. However, these results are derived from models that assume the same increase in expertise for each individual, and the data from these studies does not include repeated measures of individual success through time. In contrast, our analyses revealed that there was no population increase in expertise; instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition. For instance, many predators increased their success (Appendix S1: Figure S1A), but others displayed a decrease in success (Appendix S1: Figure S1B) or no change in success at all with experience. We predicted that this would occur because predators should differ in the type of prey that they encounter, as some may be easier to capture than others. This is often the case in nature as prey can use variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. However, prey speed in our study did not shape individual differences in expertise acquisition by predators, even if predators differed in the prey that they encountered. This was unexpected, because prey speed is an important predictor of predator success in this system as well as in others [@Walker.etal2005;@Kelley.Magurran2011;@FraserFranco.etal2022]. Despite that other antipredator tactics may mediate these individual responses by predators, it is likely that individual differences in motivation and perseverance, attention span, or memory retention were also major causes [@Warburton2003;@morand-ferronStudyingEvolutionaryEcology2016;@Dukas2019]. For instance, a persevering predator may increase its success much more steadily than an unmotivated predator that quickly abandons a chase. Another factor that may explain differences in the development of expertise is the time delay between hunting events. For example, a predator that played 300 matches in the span of six months might forget more critical information (e.g. prey escape patterns or muscle memory) than one that played 300 matches in the span of 6 days. While this has not been formally tested, a greater delay between exposure events is hypothesized to attenuate or even negate the relationship between cumulative experience and success [@Endler1991;@Wright.etal2022].

@@

The predator population displayed an increase in foraging flexibility when reaching the intermediate level of experience. As it reached an advanced level of experience, it returned to a level of flexibility that was similar to novice levels. This is commonly observed in juvenile predators across the animal kingdom, because exploring and learning different tactics at this developmental stage is crucial to become a skillful hunter [@Vehanen2003;@Johnson.Wilbrecht2011;@Thiers.etal2014;@deGrissac.etal2016]. Thus, the predators in *DBD* were probably exploring and refining different tactics at this stage of their development in the game. Interestingly, individuals became increasingly distinct in their level of foraging specialization with experience, which was related to the degree of variability in prey encounters. Predators that experienced predictable encounters with their prey specialized in hunting at high speeds, while those that experienced less predictable encounters adopted a slower and flexible strategy. This is similar to other studies finding that fast-paced hunting is a highly specialized tactic suited for prey that use rapid escapes [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018;@Szopa-Comley.Ioannou2022]. On the contrary, instead of focusing solely on one tactic, predators that experienced unpredictable encounters probably adjusted their behaviour to minimize the consequences of uncertainty. Importantly, 13% of the predator population displayed the greatest (i.e. > than 0.2 standard deviations) change in foraging specialization/flexibility with experience, while 43% remained relatively stable, and the remaining 56% displayed lower changes. Thus, while individual differences in specialization did increase with experience, they were in part driven by a proportion of the population that displayed greater changes in behaviour compared to others.

As they gained experience, predators increasingly matched their tactic to their prey, suggesting that they learned how to hunt via repeated interactions. However, while the mean speed of predators was strongly and increasingly matched to the mean speed of the prey, further investigations revealed that this apparent increase was driven largely by one individual. Yet, predators also adjusted their degree of flexibility to their prey even if the change was not very strong (Figure 3 and Table S1). In that regard, despite that only a portion of the population displayed large changes in specialization, and that adjustments to the prey were not as strong as we expected, our results still provide direct evidence that the mechanism underlying the resource-predictability hypothesis may indeed involve predators learning to adjust their degree of specialization based on their prey [@Weimerskirch.etal2005;@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017]. This is further supported by the fact that specialist and flexible hunters both achieved similar success throughout the study period.

An emerging pattern reported in predator-prey research is that predators and prey often match their phenotype to one another (i.e. positive trait covariances), which appears to be driven by an arms-race-like reciprocal phenotypic plasticity [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@Mougi.etal2011;@McGhee.etal2013]. From an evolutionary perspective, @BrodieIII.BrodieJr.1999 showed in the garter snake (*Thamnophis sirtalis*) - roughskin newt (*Taricha granulosa*) system that such escalations may occur when prey develop defensive traits, which leads to stronger selection in predators, resulting in counteradaptations that circumvent the prey's defenses. In our system, the predator's success is negatively correlated with the prey's speed, which is an effective antipredator tactic [@FraserFranco.etal2022]. However, the current study showed that as the covariance between predator speed and prey speed increased with experience, reminiscent of an arms race, we found that the consequences of hunting faster prey decreased with experience. Moreover, this pattern was followed by a decrease in among individual variation in success with experience. This has important implications for the role of experience on predator-prey interactions, as we found that learning enables predators to effectively develop countermeasures to the prey's defenses within their lifetime (i.e. at nonevolutionary timescales). It also provides empirical evidence for the hypothesis that reciprocal phenotypic plasticity through learning may attenuate selection exerted by prey on predators [@Anderson1995;@Ancel1999;@Borenstein.etal2006;@Paenke.etal2007].

@@

The interactions of predator and prey traits are probably the most important processes involved in population cycles and are expected to occur when predators match their phenotype to their prey [@abramsEvolutionPredatorPreyInteractions2000]. Yet, its has remained largely unknown whether this matching results from predators learning how to capture their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where we could directly monitor interactions, we found that experience drives individual specialization, and that predators adjust their behaviour to their prey. Moreover, individuals displayed large differences in the development of their expertise, which could result from differences in learning, as we found that almost half the population did not change their behaviour with experience. Our results have implications for predator-prey models of trait-matching, because such models are often studied at evolutionary timescales. Therefore, future studies should incorporate individual variation in learning and prey encounters to better predict under which ecological contexts specialization should be favoured over flexibility. Lastly, virtual systems are increasingly recognized among ecologists as effective systems to test hypotheses on consumer-resource interactions [@Barbe.etal2020;@Beauchamp2020;@Cere.etal2021;@FraserFranco.etal2022], but also in other fields such as citizen science and conservation [@Sandbrook.etal2015;@Redpath.etal2018;@Duthie.etal2021]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

@@

>>>>

Hunting experience is predicted to increase predator foraging specialization when resources are predictable [@Estes.etal2003;@Tinker.etal2008;@Tinker.etal2009]. However, when resources vary, others predict that learning with experience should increase behavioural flexibility [@Stephens1993;@Ishii.Shimada2010;@Mery.Burns2010;@Snell-Rood2013;@Wright.etal2022]. Our results show that both mechanisms operated simultaneously, such that there was no directional increase in behavioural specialization or flexibility in the population. Instead, the consequences from experience appear to be a property of individuals, such that each predator developed, through extensive practice, their own expertise and behavioural trajectory; some specialized at fast-paced hunting, while others adopted a slower and more flexible tactic. The predators' tactics were matched to prey encounters, such that specialized cursorial hunters experienced more predictable encounters, while slower and flexible hunters experienced more unpredictable encounters, with no difference in success between the two predator-types. This predator-prey behavioural matching grew stronger with predator experience, while the relationship between the prey's behaviour and hunting success diminished. Together, our observations suggest that predators adjust their hunting tactic with experience, and the degree at which they specialize in it, to the behaviour of the prey that they encountered throughout their lifetime in the game. Moreover, the predator-prey behavioural escalation through experience suggests that learning may be a catalyst for arms races.

@@

Empirical research in humans and animals shows that task proficiency often increases nonlinearly with experience and stabilizes at an expert level [reviewed in @Dukas2019]. However, we cannot derive conclusions at the individual-level from these studies either because their models assume the same increase in expertise for each individual, or because it is unclear whether their data includes repeated measures of individual success through time. In contrast, our analyses revealed that there was no population increase in expertise; instead, individual predators in *DBD* displayed distinct patterns of expertise acquisition. For instance, many predators increased their success (Appendix S1: Figure S1A), but others displayed a decrease in success (Appendix S1: Figure S1B) or no change in success at all with experience. We predicted that this would occur because predators should differ in the type of prey that they encounter, as some may be easier to capture than others. This is often the case in nature as prey can use variety of defenses such as physical armaments, toxins, camouflage, or antipredator behaviours to escape predation [@BrodieIII.BrodieJr.1999;@Bowen.etal2002;@brodieCoEvolutionPredatorsPrey2010;@Nomura.etal2011;@Carey.Wahl2011]. However, prey speed in our study did not shape individual differences in expertise acquisition by predators, even if predators differed in the prey that they encountered. This was unexpected, because prey speed is an important predictor of predator success in this system as well as in others [@Walker.etal2005;@Kelley.Magurran2011;@FraserFranco.etal2022]. Despite that other antipredator tactics may mediate these individual responses by predators, it is likely that individual differences in motivation and perseverance, attention span, or memory retention were also major causes [@Warburton2003;@morand-ferronStudyingEvolutionaryEcology2016;@Dukas2019]. For instance, a persevering predator may increase its success much more steadily than an unmotivated predator that quickly abandons a chase. Another factor that may explain differences in the development of expertise is the time delay between hunting events. For example, a predator that played 300 matches in the span of six months might forget more critical information (e.g. prey escape patterns or muscle memory) than one that played 300 matches in the span of 6 days. While this has not been formally tested, a greater delay between exposure events is hypothesized to attenuate or even negate the relationship between success and experience [@Endler1991;@Wright.etal2022].

@@

The predator population displayed an increase in foraging flexibility when reaching the intermediate level of experience. As it reached an advanced level of experience, it returned to a level of flexibility that was similar to novice levels. This is commonly observed in juvenile predators across the animal kingdom, because exploring and learning different tactics at this developmental stage is crucial to become a skillful hunter [@Vehanen2003;@Johnson.Wilbrecht2011;@Thiers.etal2014;@deGrissac.etal2016]. Thus, the predators in *DBD* were probably exploring and refining different tactics at this stage of experience. Interestingly, individuals became increasingly distinct in their level of foraging specialization with experience, which was related to the degree of variability in prey encounters. Predators that experienced predictable encounters with their prey specialized in hunting at high speeds, while those that experienced less predictable encounters adopted a slower and flexible strategy. This is similar to other studies finding that fast-paced hunting is a highly specialized tactic suited for prey that use rapid escapes [@Endler1991;@Bro-Jorgensen2013;@Wilson.etal2018;@Szopa-Comley.Ioannou2022]. On the contrary, instead of focusing solely on one tactic, predators that experienced unpredictable encounters probably adjusted their behaviour to minimize the consequences of uncertainty. Importantly, 13% of the predator population displayed the greatest change in foraging specialization/flexibility with experience (i.e. $\ge$ 0.2 change in standard deviation), 56% displayed lower changes (i.e. >0.05 and <0.2 change in standard deviation), while 43% remained relatively stable (i.e. $\le$ 0.05 change in standard deviation). Thus, while individual differences in specialization did increase with experience, they were in part driven by a proportion of the population that displayed greater changes in behaviour compared to others.

As they gained experience, predators increasingly matched their tactic to their prey, suggesting that they learned how to hunt via repeated interactions. However, while the mean speed of predators was strongly and increasingly matched to the mean speed of the prey, further investigations revealed that this apparent increase was driven largely by one individual. Yet, predators also adjusted their degree of flexibility to their prey even if the change was not very strong (Figure 3 and Appendix S2: Table S1). In that regard, despite that only a portion of the population displayed large changes in specialization, and that adjustments to the prey were not as strong as we expected, our results still provide direct evidence that the mechanism underlying the resource-predictability hypothesis may indeed involve predators learning to adjust their degree of specialization based on their prey [@Weimerskirch.etal2005;@Weimerskirch2007;@Woo.etal2008;@Phillips.etal2017]. This is further supported by the fact that specialist and flexible hunters both achieved similar success throughout the study period.

An emerging pattern reported in predator-prey research is that predators and prey often match their phenotype to one another (i.e. positive trait covariances), which appears to be driven by an arms-race-like reciprocal phenotypic plasticity [@Kishida.etal2006;@Kishida.etal2009;@Edgell.Rochette2009a;@Mougi.etal2011;@McGhee.etal2013]. From an evolutionary perspective, @BrodieIII.BrodieJr.1999 showed in the garter snake (*Thamnophis sirtalis*) - roughskin newt (*Taricha granulosa*) system that such escalations may occur when prey develop defensive traits, which leads to stronger selection in predators, resulting in counteradaptations that circumvent the prey's defenses. In our system, the predator's success is negatively correlated with the prey's speed, which is an effective antipredator tactic [@FraserFranco.etal2022]. However, we found that as the covariance between predator speed and prey speed increased with experience, reminiscent of an arms race, the consequences of hunting faster prey decreased with experience. Moreover, this pattern was followed by a decrease in among individual variation in success with experience. This has important implications for the role of experience on predator-prey interactions, as our results suggest that learning enables predators to effectively develop countermeasures to the prey's defenses within their lifetime (i.e. at nonevolutionary timescales). It also provides empirical evidence for the hypothesis that reciprocal phenotypic plasticity through learning may attenuate selection exerted by prey on predators [@Anderson1995;@Ancel1999;@Borenstein.etal2006;@Paenke.etal2007].

@@

The interactions of predator and prey traits are probably the most important processes involved in population cycles and are expected to occur when predators match their phenotype to their prey [@abramsEvolutionPredatorPreyInteractions2000]. Yet, its has remained largely unknown whether this trait-matching results from predators learning how to capture their prey, in part because of the challenges of investigating direct interactions in the wild. By capitalizing on a virtual predator-prey system where we could directly monitor interactions, we found that experience drives individual specialization, and that predators adjust their behaviour to their prey. Moreover, individuals displayed large differences in the development of their expertise, which could result from differences in learning, as we found that almost half the population did not change their behaviour with experience. Our results have implications for predator-prey models of trait-matching, because such models are often studied at evolutionary timescales. Therefore, future studies should incorporate individual variation in experience to better predict under which ecological contexts specialization should be favoured over flexibility. Lastly, virtual systems are increasingly recognized among ecologists as effective systems to test hypotheses on consumer-resource interactions [@Barbe.etal2020;@Beauchamp2020;@Cere.etal2021;@FraserFranco.etal2022], but also in other fields such as citizen science and conservation [@Sandbrook.etal2015;@Redpath.etal2018;@Duthie.etal2021]. We therefore hope that our study will inspire more collaborations between scientists and the videogame industry to tackle fundamental questions in ecology.

@@

Discussion :

- peut-être ajouter le fait que malgré que la vitesse soit mieux pour les proies, elles pourraient buffer ça en étant + imprévisibles.

- peut être ajouter que l'escalade est aussi dû au fait que la proie pourrait être plus susceptible à la prédation en étant rapide car elle s'expose plus.
